{\bf \Large TO BE RENAMED}

### 2023-06-01 draft.tex
finished rework, rewritten the conclusion, corrected grammatical mistakes

\usepackage[activeacute,english]{babel}
%\usepackage{fancyhdr}
\usepackage{mathrsfs}
\usepackage{authblk}
\usepackage[alwaysadjust]{paralist}
\usepackage{alltt}
\usepackage{caption}
\usepackage{array}
\usepackage[fit]{truncate}
%\pagestyle{fancy}
\usepackage{calc}
%\usepackage{fancyvrb}
\usepackage{float}
\usepackage[ansinew]{inputenc}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage[figuresright]{rotating}
\usepackage{color}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usetikzlibrary{positioning, arrows}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{array}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{array}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{adjustbox}

@@

    This thesis comprises of four studies investigating the context effects arising from multidimensional, multi-attribute choice settings and applies the knowledge to enhance recommender systems in online marketplaces. The findings indicate that context effects in choice settings which were primarily studied and observed in experimental settings before are also present and detectable in field settings. Moreover, tools that enable users to signal their preferences to recommender systems can significantly increase its adoption. The empirical tests of these chapters rely on four distinct and unique datasets, the largest being an observational data from a field setting and the remaining three coming from an experimental setting.

@@

    The third study presented a novel approach to addressing the user-side cold-start problem in recommender system design. It applies the findings of two-stage choice process observed in individuals to contextual information extracted from the choice set via clustering to generate consideration sets. The approach proves to be robust to clustering method and different heuristics applied. Its findings pave the way for investigating context effects arising from outside the choice sets, namely the preferences of individuals and the tools that allow them to signal their preferences to recommender systems. This is argued to be one of the main mechanisms to create more effective recommender systems.

@@

    This thesis demonstrates the existence of context effects in multi-attribute, multidimensional settings and develops methodologies on enhancing recommender systems design with context effects. Also, this study investigates context arising outside the choice set, namely, the tools that allows users to express their preferences and how to make recommender systems better through them.

@@

\section{General introduction}

@@

Recommender systems have become a crucial ally in the vast landscape of digital world. They help individuals to navigate through the plethora of alternatives and find what they are looking for. They achieve this by using sophisticated algorithms that analyze the wide space of items, users and their interactions with each other. System designers and businesses thrive to maximize the accuracy of recommendations, meaning they want to maximize the consumption of the recommendations. However, recent research stream shows that, accuracy is not the only way to make recommender systems effective. Other factors like diversity, serendipity \citep{kaminskas2016diversity}. Moreover, the black-box nature of the algorithms do not allow to understand the reasons behind a particular user's behaviour \citep{samih2021exmrec2vec, kotkovSurveySerendipityRecommender2016}. Hence, more and more interest is driven towards the understanding and utilizing the choice context \citep{adomavicius2005toward}. Choice context has received interest in scenarios where user or item information is scarce \citep{scheinMethodsMetricsColdStart}.

Choice context has long received the attention of scholars in many fields, including marketing, psychology, economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates and uses two of them. The first one posits that choice context \footnote{For the remainder of the paper, I will follow the previous literature \citep{truebloodEtAl13} and refer this definition of choice context as context effects} is "the availability and the nature of choice alternatives" \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Context also referred to as "the time and content of the choice, the location or socio-demographic characteristics of the decision-maker..." \citep{adomavicius2011context}.

@@

When new users or items are introduced to recommender systems, they fail to function as intended because of the lack of information their algorithms need. In such cases, even when they do provide recommendations, they are far from personalized \citep{lika2014facing}. It is considered to be a key challenge in recommender system design \citep{park2009pairwise}. Previous literature has applied various methods to overcome this problem \citep{aharon2013off, bykau2013coping, saveski2014item, guy2009personalized}. However, the limitations of these approaches are that they ignore the context effects and concentrate on scenarios where information scarcity is temporary. Using the findings from my second study, decision-making and choice modeling literature I propose an innovative solution to address continious cold-start issue in recommender system design by utilizing menu contextual information to generate relevant choice sets through a two-step choice modeling method.

Research agrees on importance of metrics beyond accuracy for recommender systems \citep{kaminskas2016diversity}. It is in the best interest of online marketplaces to provide users with recommendations which are not only in the category "exactly what I want", but also "I never thought I would have liked this one" \citep{kotkovSurveySerendipityRecommender2016}. This can not only boost sales \citep{songWhenHowDiversify2019}, but also increase the satisfaction of users \citep{knijnenburgExplainingUserExperience2012}. Albeit, it was observed that the preferences of individuals are not stable and tend to change and system designers have proposed various tools to enable users signal their preference shifts \citep{bostandjiev2012tasteweights,hijikata2012relation}. However, to better understand it is necessary to study those tools in isolation from the context arising from the choice set and concentrate purely on context arising from the user side, e.g their preferences \citep{adomavicius2011context}. In my last study I address this gap by conducting an online experiment and applying Technology Acceptance Model \citep{davis1985technology} to measure users' acceptance of such tools.

@@

\section{Application of mathematical decision making model to observational data} \footnote{This chapter is based on a joint work with my supervisor  Zakaria Babutsidze}

@@

Context effects have been extensively studied and demonstrated in various domains, from psychology to marketing \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  truebloodEtAl13, wuConsguner20, frederickEtAl14}. Some recent studies also concluded that multiple context effects may occur at the same time \citep{berkowitsch2014rigorously, noguchi2014attraction}. However, recent studies also have discovered boundary conditions for these effects \citep{liew2016appropriacy, spektor2018good, spektor2019similarity}. Familiarity with the choice domain was found to reduce context effects experienced by individuals \citep{kim2005attraction, sheng2005understanding}.  It was also found that, some conditions may force these effects to completely reverse \citep{cataldo2019comparison}. The findings described above make it necessary to call for a model which could explain these effects.

@@

In the last three decades, researchers have developed many computational models which account for context effects: Multialternative Decision Field Theory (MDFT) \citep{roe2001multialternative}, Leaky competing
accumulator \citep{usher2001time}, Multiattribute linear ballistic accumulator \citep{trueblood2014multiattribute}, Multialternative Decision by Sampling  \citep{noguchi2018multialternative} and etc.. Some of these models have been extensively tested and studied, while others are relatively new, hence have not received much attention from scholars \citep{truebloodEtAl13}. However, these studies have been performed with experimental data \citep{evans2019response, berkowitsch2014rigorously, trueblood2014multiattribute, busemeyer2019cognitive}. Research has proven that the behavior of individuals in laboratory choice environment is different from real-world choice environment \citep{hogarth1989risk}. Hence, the applicability of such models to field data is an uninvestigated avenue, because no previous study has been done where a computational model was been applied to a real-world observational data. I plan to address this gap by applying Multialternative Decision by Sampling (MDbS) model proposed by Noguchi and Stewart \citeyearonly{noguchi2018multialternative} to a field data from airfare booking domain. This would allow me to assess the applicability level of this model to multialternative and multidimensional field data.

Also, applying MDbS to observational data would allow me to statistically asses the significance of the contribution of this model's ability to account for context effect. Instead of testing this model against established choice models, I will attempt to complement them with MDbS. To do this, I will use random effect Probit model as a variation of Probit family and augment it with MDbS output. I choose Probit model as it does not explictly assume IIA unlike family of logit models. To validate my results further I will apply the same methodology and analysis to an experimental data.

I have chosen MDbS for two main reasons. Firstly, it is relatively new when compared to other models, hence it has not been further investigated before. Secondly, it is more robust and can account for wider range of context effects (than other models) known to the literature \citep{noguchi2018multialternative}. I will discuss these reasons in more detail in the next section.

@@

If this situation is familiar to you, then you have experienced so called context effect which can bu understood as "the composition  and the nature of the choice set, availability of various options in it" \citep{tversky1972elimination, huberPuto83}. For the long time, our understanding of the choice did not extend over the borders of two related principles. First one is, independence of irrelevant alternatives (IIA) which states that when having a choice between two two options $A$ and $B$ if a person prefers $A$ for example, regardless of adding a third option $C$ to this choice set, that person's preference must be unaltered \citep{luce59}. Second one is regularity principle which states that, the choice probability of option $A$ can not increase by introduction of option $C$ \citep{luce59}.

However, research has concluded that the context of the choice set and options in it have substantial effect on how people make choices. This effect has been studied extensively for the past five decades by many economists, marketing scholars and psychologists \footnote{See Dowling et al. \citeyearonly{dowlingEtAl20} and Lichtenstein \citeyearonly{lichtenstein2006construction} for more comprehensive review.} \citep{tverskySimonson93, kahnemanTversky79, simonson89, lichtenstein2006construction, dowlingEtAl20}. Most of the research has focused on three context effects, also reffered to as "big three": attraction, compromise and similarity \citep{howes2016contextual}.

@@

People have heterogeneous preferences and hence when facing a choice set with only these two options they would weigh each dimension roughly equally. The ones who prefer Dimension 1 would choose $B$ whereas $A$ will be chosen by individuals preferring Dimension 2. This can be described as equation \ref{eq:onlyTwoOptions} below.

@@

Now, lets add a third option to this choice set, option $D_A$ to create one variation and $D_B$ to create the second variation of the ternary choice set. Both added options have lower values in both dimensions when compared to $A$ and $B$ respectively. Huber created this type of scenario and has found what he has then called "attraction effect" \citeyearonly{huberEtAl82}. Attraction effect, which is also known in the literature as asymmetric dominance effect is a consistent violation of the regularity principle mentioned earlier. He suggested that when having a choice set consisting of options $A$ and $B$ the relative probability of choosing option $A$ can be increased if one adds a third option with characteristics of $D_A$ to the same choice set \citeyearonly{huberEtAl82}.

@@

It is worth noting that, it is also possible to "target" a particular option from binary choice set when adding a third option to create a compromise effect. One can add a target $C$ which makes $A$ a compromise option. In this case, the probability of choosing $A$ among this triple will increase, as it is considered a compromise between the remaining two options.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{staticFiles/compromiseEffect.png}
    \caption{Compromise effect in ternary choice. ALSO TALK ABOUT MAKING ANY OF THE ORIGINAL POINTS COMPRMISE BY ADDING THIRD OPTION} % Add your description here
    \label{fig:compromiseEffect} % This labels your figure for reference

@@

Although Becker et al. \citeyearonly{becker1964measuring} have mentioned it before, the first study of similarity effect is known to be the one by Tversky \citep{tversky1972elimination}. He noted that when facing a binary choice set consisting of $A$ and $B$ individuals will gravitate towards $A$ more than when facing a ternary choice set consisting of ${A, B, S_A}$ depicted in figure \ref{fig:similarityEffect}. He explained it by proposing elimination by aspects theory, which states that one attribute will be chosen as elimination criteria, and all options that do not meet that criteria will be eliminated \citep{tversky1972elimination}. Hence in the choice set ${A, B, S_A}$ if an individual selects dimension 2 as elimination criteria, both $A$ and $S_A$ will be eliminated, leaving $B$ as a choice. On the contrary, if decision maker prefers dimension 1 more, then option 
$B$ will be eliminated, leaving both $A$ and $S_A$ to share the "victory", hence resulting in equation \ref{eq:similarityProbability}.

@@

MDbS belongs to the attention based choice models. Before commencing with its underlying mechanisms and assumptions, it is beneficial to discuss other two models which are preceding MDbS and have been studied extensively. After briefly discussing those models I will continue with MDbS, its main assumptions, mechanisms and how it accounts

@@

MDbS has its origins in theory of decision by sampling which assumes that individual preferences arise from binary, ordinal comparisons of alternatives on given attribute values with reference values from the memory \citep{stewart2006decision}. Unlike it, MDbS assumes that the information required for comparison also comes from the choice environment itself \citep{noguchi2018multialternative}. As the other two previous models discussed above, its mechanisms can be explained using three stages. To better understand these stages I will discuss them in detail.

@@

According to MDbS, when comparing two tickets between Paris and New-York, the price of a ticket would be compared to the prices of other tickets in the choice set and also to the ones and which an individual has previously seen, but are not in the current choice set. Comparisons are ordinal, meaning that evidence accumulated towards the "winner" at a rate of one irregardless of how large the difference was.

@@

\begin{align}\label{similarityMDBS}
s_{ij,kj} = \exp \left( - \alpha \left| \frac{m_{ij} - m_{kj}}{m_{kj}} \right| \right) 
\end{align}

with similarity parameter $\alpha$. Also, generally $s_{ij,kj} \neq s_{kj,ij}.$ Consider also

\begin{align}\label{sumOfSimilaritiesMDbS}
    s_{ij} = \sum_{\substack{k \neq l \in \{1, \ldots, n_a\}}} s_{ij,kj}
\end{align}
to be the sum of all similarities for attribute $m_{ij}$ to other attributes on the same dimension. Consequently, by dividing this value to the sum of similarities across all other attributes across all dimensions, one can calculate the probability that $m_{ij}$ will be selected for comparison which will be

@@

\begin{align}\label{probabilityOneIsFavored}
    P(m_{ij} \text{ is favored over } m_{kj}) = 
        \begin{cases}
        F(\beta_0 (| \frac{m_{ij} - m_{kj}}{m_{kj}} |- \beta_1)) & \text{if } A_i > X_i \\
        0 & \text{otherwise}
        \end{cases}
\end{align}

where $F$ is a logistic sigmoid function and $\beta_0$ and $\beta_1$ corresponde to the advantage value and the probability that this particular advantage value will be enough to be preferred. For example, consider the case where $\beta_0 = 0.1$ and $\beta_1 = 50$. This would mean that the advantage of 10\% would be preferred with 50\% probability. Consequently, if the difference is 20\%, then it will be preferred with 99\% probability. The logistic function brings the notion of "soft" comparison instead of "hard" comparison in which case the small differences would be ignored, while large differences would be extremely preferred \citep{noguchi2018multialternative}.

@@

I will begin this section with introducing an example dataset depicted in figure \ref{fig:MDBsContextExample} which I intend to use when explaining how MDbS accounts for a given context effect. Although there are five alternatives in the figure \ref{fig:MDBsContextExample}, only three of them will be discussed at a time.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{staticFiles/contextEffectExampleScatterplot.png}
    \caption{Example choice set to explain MDbS' account for big three context effects. $A$ and $B$ are considered original two alternatives (binary choice set). $D$ is dominated by $A$ on both dimensions. $C$ acts as compromise between the original two and $S$ is a similar option to $B$. Although these effects would be present in different variations of the choice set (for example one can make $A$ as compromise option), for the simplicity I will concnetrate on this example.} % Add your description here
    \label{fig:MDBsContextExample} % This labels your figure for reference

@@

When adding option $D$ which is dominated by $A$ in both dimensions to binary choice set $A$ $B$ (hereafter binary choice set will be used instead of $A$ and $B$) one creates an attraction effect \citep{huberEtAl82, huberPuto83}. Huber et al. \citeyearonly{huberEtAl82} explained attraction effect via weight shifts for individuals. Addition of $D$ would make people to weigh dimension 2 more. Hence, $A$ and $D$ will have higher "interest" among people, and this is where $A$ wins over $D$.

@@

In MDbS this is explained by $\beta_0$ and $\beta_1$ from equation \ref{probabilityOneIsFavored}. Recall that, the reason of having the sigmoid function with arguments $\beta_0$ and $\beta_1$ is to make sure that small differences would be relatively ignored. This is in line with previous literature stating that people tend to ignore small differences between alternatives when making a choice \citep{kalwani1992consumer}. Hence, the small differences between $B$ and $S$ would be ignored, which would be translated to decreased probability of any of them winning over the other when being compared. This will indirectly increase the evidence accumulation for $A$ while resulting in "shared" evidence between $B$ and $S$.

Not only MDbS accounts for big three context effect, it also successfully accounts for other known effects to decision making literature such as: attribute spacing effect \citep{cooke1998multiattribute}, centrality effect \citep{brown2011decision}, background contrast effect \citep{tverskySimonson93}, endowment effect \citep{knetsch1989endowment} and etc \footnote{For the full list of the effects MDbS can account for please refer to Noguchi \citeyearonly{noguchi2018multialternative}}. Overall, authors claim that MDbS can theoretically account for up to 25 context effect variations.

@@

Having discussed the theoretical aspects of MDbS and its account for big three context effects, I know proceed to the discussion of practical aspects of the research: the methodology. In the next section, I delve into the specifics of the research process where I briefly discuss the dataset and provide detailed account of the steps taken in the analysis.

@@

This section will discuss the methodology I have used for this study. After brief discussion of the dataset I will move onto detailed discussion of the steps taken in the analysis.

@@

Attribute range effect which was first investigated by Mellers \citeyearonly{mellers1994trade} which described people's tendency to scale perceived attractiveness of an alternative on given attribute using the entire range of that attribute. Hence, $\beta_0$ in MDbS represents the fraction of the difference between attributes compared to the entire attribute range which is bounded between 0 and 1. On the other hand, $\beta_1$ represents the percentage of preference of that difference for an individual, hence it is also bounded with values between 0 and 100. I have created 99 $\beta_0$ values evenly spaced between 0 and 1 and 99 $beta_1$ values evenly spaced between 0 and 100.

@@

\begin{algorithm}
\caption{Simplified Differential Evolution}
\begin{algorithmic}[1]

@@

 $$\text{Similarity}(F) = O(\text{Cluster}_F)
$$

>>>>

\usepackage[activeacute,english]{babel}
%\usepackage{fancyhdr}
\usepackage{mathrsfs}
\usepackage{authblk}
\usepackage[alwaysadjust]{paralist}
\usepackage{alltt}
\usepackage{caption}
\usepackage{array}
\usepackage[fit]{truncate}
%\pagestyle{fancy}
\usepackage{calc}
%\usepackage{fancyvrb}
\usepackage{float}
\usepackage[ansinew]{inputenc}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage[figuresright]{rotating}
\usepackage{color}
\usepackage{hyperref}
\usepackage{enumerate}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usetikzlibrary{positioning, arrows}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{array}
\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{array}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{adjustbox}
\usepackage[stable]{footmisc}

@@

    This dissertation comprises of four studies investigating the context effects and applies the knowledge to enhance recommender systems in online marketplaces. I differentiate between two types of context. The first type is intricately tied to the alternatives present within a choice set. The second type of context originates from factors external to the choice set. The findings of the dissertation indicate that internal context effects in choice settings which were primarily studied and observed in experimental settings before are also present and detectable in field settings. Moreover, the study investigating the external context finds the positive relationship between tools that enable users to signal their preferences to recommender systems and their adoption. The empirical applications of the chapters comprising this thesis rely on four distinct and unique datasets, the largest being an observational data from a field setting and the remaining three coming from an experimental setting.

@@

    The third study presents a novel approach to addressing the user-side cold-start problem in recommender system design. It applies the findings of two-stage choice process observed in individuals to contextual information extracted from the choice set via clustering to generate consideration sets. The approach proves to be robust to clustering method and different heuristics applied. Its findings pave the way for investigating context effects arising from outside the choice sets, namely the preferences of individuals and the tools that allow them to signal their preferences to recommender systems. This is argued to be one of the main mechanisms to create more effective recommender systems.

@@

    This thesis demonstrates the existence of context effects in multi-attribute, multidimensional settings and develops methodologies on enhancing recommender systems design with context effects. Additionally, this study investigates external context, namely, the tools that allows users to express their preferences and how to make recommender systems better through them.

@@

\section{Introduction}

@@

Recommender systems have become a crucial ally in the vast landscape of digital world. They help individuals to navigate through the plethora of alternatives and find what they are looking for. They achieve this by using sophisticated algorithms that analyze the wide space of items, users and their interactions with each other. System designers and businesses thrive to maximize the accuracy of recommendations, meaning they want to maximize the consumption of the recommendations. However, recent research stream shows that, accuracy is not the only way to make recommender systems effective. Other factors like diversity, serendipity \citep{kaminskas2016diversity}. Moreover, the black-box nature of the algorithms do not allow to understand the reasons behind a particular user's behaviour \citep{kotkovSurveySerendipityRecommender2016, samih2021exmrec2vec}. Hence, more and more interest is driven towards the understanding and utilizing the choice context \citep{adomavicius2005toward}. Choice context has received interest in scenarios where user or item information is scarce \citep{scheinMethodsMetricsColdStart}.

Choice context has long received the attention of scholars in many fields, including marketing, psychology, economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates and uses two of them. The first one posits that choice context \footnote{For the remainder of the paper, I will follow the previous literature \citep{truebloodEtAl13} and refer this definition of choice context as context effects.} is ``the availability and the nature of choice alternatives" \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Context also referred to as ``the time and content of the choice, the location or socio-demographic characteristics of the decision-maker..." \citep{adomavicius2011context}.

@@

When new users or items are introduced to recommender systems, they fail to function as intended because of the lack of information their algorithms need. In such cases, even when they do provide recommendations, they are far from personalized \citep{lika2014facing}. It is considered to be a key challenge in recommender system design \citep{park2009pairwise}. Previous literature has applied various methods to overcome this problem \citep{guy2009personalized, aharon2013off, bykau2013coping, saveski2014item}. However, the limitations of these approaches are that they ignore the context effects and concentrate on scenarios where information scarcity is temporary. Using the findings from my second study, decision-making and choice modeling literature I propose an innovative solution to address continious cold-start issue in recommender system design by utilizing menu contextual information to generate relevant choice sets through a two-step choice modeling method.

Research agrees on importance of metrics beyond accuracy for recommender systems \citep{kaminskas2016diversity}. It is in the best interest of online marketplaces to provide users with recommendations which are not only in the category ``exactly what I want", but also ``I never thought I would have liked this one" \citep{kotkovSurveySerendipityRecommender2016}. This can not only boost sales \citep{songWhenHowDiversify2019}, but also increase the satisfaction of users \citep{knijnenburgExplainingUserExperience2012}. Albeit, it was observed that the preferences of individuals are not stable and tend to change and system designers have proposed various tools to enable users signal their preference shifts \citep{bostandjiev2012tasteweights,hijikata2012relation}. However, to better understand it is necessary to study those tools in isolation from the context arising from the choice set and concentrate purely on context arising from the user side, e.g their preferences \citep{adomavicius2011context}. In my last study I address this gap by conducting an online experiment and applying Technology Acceptance Model \citep{davis1985technology} to measure users' acceptance of such tools.

@@

\section{Application of mathematical decision making model to observational data}\label{chapter:simulationStudy}
\begin{abstract}

@@

Context effects have been extensively studied and demonstrated in various domains, from psychology to marketing \citep{herne1997decoy, soltani2012range, truebloodEtAl13, frederickEtAl14, evangelidisEtAl18, wuConsguner20}. Some recent studies also concluded that multiple context effects may occur at the same time \citep{berkowitsch2014rigorously, noguchi2014attraction}. However, recent studies also have discovered boundary conditions for these effects \citep{liew2016appropriacy, spektor2018good, spektor2019similarity}. Familiarity with the choice domain was found to reduce context effects experienced by individuals \citep{kim2005attraction, sheng2005understanding}.  It was also found that, some conditions may force these effects to completely reverse \citep{cataldo2019comparison}. The findings described above make it necessary to call for a model which could explain these effects.

@@

In the last three decades, researchers have developed many computational models which account for context effects: Multialternative Decision Field Theory (MDFT) \citep{roe2001multialternative}, Leaky competing
accumulator \citep{usher2001time}, Multiattribute linear ballistic accumulator \citep{trueblood2014multiattribute}, Multialternative Decision by Sampling  \citep{noguchi2018multialternative}. Some of these models have been extensively tested and studied, while others are relatively new, hence have not received much attention from scholars \citep{truebloodEtAl13}. However, these studies have been performed with experimental data \citep{trueblood2014multiattribute,  berkowitsch2014rigorously, evans2019response, busemeyer2019cognitive}. Research has proven that the behavior of individuals in laboratory choice environment is different from real-world choice environment \citep{hogarth1989risk}. Hence, the applicability of such models to field data is an uninvestigated avenue, because no previous study has been done where a computational model was been applied to a real-world observational data. The plan is to address this gap by applying Multialternative Decision by Sampling (MDbS) model proposed by Noguchi and Stewart \citeyearonly{noguchi2018multialternative} to a field data from airfare booking domain. This would allow me to assess the applicability level of this model to multialternative and multidimensional field data.

Also, applying MDbS to observational data would allow me to statistically asses the significance of the contribution of this model's ability to account for context effect. Instead of testing this model against established choice models, I will attempt to complement them with MDbS. To do this, random effect Probit model as a variation of Probit family will be used and augmented with MDbS output. Thee Probit model is chosen as it does not explictly assume IIA unlike family of logit models. To validate my results further I will apply the same methodology and analysis to an experimental data.

The reasons behind choosing MDbS are twofold. Firstly, it is relatively new when compared to other models, hence it has not been further investigated before. Secondly, it is more robust and can account for wider range of context effects (than other models) known to the literature \citep{noguchi2018multialternative}.

@@

If this situation is familiar to you, then you have experienced so called context effect which can bu understood as ``the composition  and the nature of the choice set, availability of various options in it" \citep{tversky1972elimination, huberPuto83}. For the long time, our understanding of the choice did not extend over the borders of two related principles. First one is, independence of irrelevant alternatives (IIA) which states that when having a choice between two two options $A$ and $B$ if a person prefers $A$ for example, regardless of adding a third option $C$ to this choice set, that person's preference must be unaltered \citep{luce59}. Second one is regularity principle which states that, the choice probability of option $A$ can not increase by introduction of option $C$ \citep{luce59}.

However, research has concluded that the context of the choice set and options in it have substantial effect on how people make choices. This effect has been studied extensively for the past five decades by many economists, marketing scholars and psychologists \footnote{See Dowling et al. \citeyearonly{dowlingEtAl20} and Lichtenstein \citeyearonly{lichtenstein2006construction} for more comprehensive review.} \citep{ kahnemanTversky79, simonson89, tverskySimonson93, lichtenstein2006construction, dowlingEtAl20}. Most of the research has focused on three context effects, also reffered to as ``big three": attraction, compromise and similarity \citep{howes2016contextual}.

@@

People who prefer Dimension 1 would choose $B$ whereas $A$ will be chosen by individuals preferring Dimension 2. This can be described as equation \ref{eq:onlyTwoOptions} below.

@@

Now, lets add a third option to this choice set, option $D_A$ to create one variation and $D_B$ to create the second variation of the ternary choice set. Both added options have lower values in both dimensions when compared to $A$ and $B$ respectively. Huber created this type of scenario and has found what he has then called ``attraction effect" \citeyearonly{huberEtAl82}. Attraction effect, which is also known in the literature as asymmetric dominance effect is a consistent violation of the regularity principle mentioned earlier. He suggested that when having a choice set consisting of options $A$ and $B$ the relative probability of choosing option $A$ can be increased if one adds a third option with characteristics of $D_A$ to the same choice set \citeyearonly{huberEtAl82}.

@@

It is worth noting that, it is also possible to ``target" a particular option from binary choice set when adding a third option to create a compromise effect. One can add a target $C$ which makes $A$ a compromise option. In this case, the probability of choosing $A$ among this triple will increase, as it is considered a compromise between the remaining two options.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{staticFiles/compromiseEffect.png}
    \caption{Compromise effect in ternary choice.} % Add your description here
    \label{fig:compromiseEffect} % This labels your figure for reference

@@

Although Becker et al. \citeyearonly{becker1964measuring} have mentioned it before, the first study of similarity effect is known to be the one by Tversky \citep{tversky1972elimination}. He noted that when facing a binary choice set consisting of $A$ and $B$ individuals will gravitate towards $A$ more than when facing a ternary choice set consisting of ${A, B, S_A}$ depicted in figure \ref{fig:similarityEffect}. He explained it by proposing elimination by aspects theory, which states that one attribute will be chosen as elimination criteria, and all options that do not meet that criteria will be eliminated \citep{tversky1972elimination}. Hence in the choice set ${A, B, S_A}$ if an individual selects dimension 2 as elimination criteria, both $A$ and $S_A$ will be eliminated, leaving $B$ as a choice. On the contrary, if decision maker prefers dimension 1 more, then option 
$B$ will be eliminated, leaving both $A$ and $S_A$ to share the ``victory", hence resulting in equation \ref{eq:similarityProbability}.

@@

MDbS belongs to the attention based choice models. These models are type of decision-making models which take into account the attention allocation mechanisms when making decisions. They generally assume that individuals allocate attention to different attributes and the ones receiving more attention is having more impact on decision making \citep{gabaix2000boundedly}. Before commencing with its underlying mechanisms and assumptions, it is beneficial to discuss other two models which are preceding MDbS and have been studied extensively. After briefly discussing those models I will continue with MDbS, its main assumptions, mechanisms and how it accounts for context effects.

@@

MDbS has its origins in theory of decision by sampling which assumes that individual preferences arise from binary, ordinal comparisons of alternatives on given attribute values with reference values from the memory \citep{stewart2006decision}. Unlike it, MDbS assumes that the information required for comparison also comes from the choice environment itself \citep{noguchi2018multialternative}. As the other two previous models discussed above, its mechanisms can be explained using three stages. Next section will discuss this in detail.

@@

According to MDbS, when comparing two tickets between Paris and New-York, the price of a ticket would be compared to the prices of other tickets in the choice set and also to the ones and which an individual has previously seen, but are not in the current choice set. Comparisons are ordinal, meaning that evidence accumulated towards the ``winner" at a rate of one irregardless of how large the difference was.

@@

\begin{align}\label{similarityMDBS}
s_{ij,kj} = \exp \left( - \alpha \left| \frac{m_{ij} - m_{kj}}{m_{kj}} \right| \right) ,
\end{align}
with similarity parameter $\alpha$. Also, generally $s_{ij,kj} \neq s_{kj,ij}.$ Consider also

\begin{align}\label{sumOfSimilaritiesMDbS}
    s_{ij} = \sum_{\substack{k \neq l \in \{1, \ldots, n_a\}}} s_{ij,kj}  ,
\end{align}
which is the sum of all similarities for attribute $m_{ij}$ to other attributes on the same dimension. Consequently, by dividing this value to the sum of similarities across all other attributes across all dimensions, one can calculate the probability that $m_{ij}$ will be selected for comparison which will be

@@

\begin{align}\label{probabilityOneIsFavored}
    P(m_{ij} \text{ is favored over } m_{kj}) = 
        \begin{cases}
        F(\beta_0 (| \frac{m_{ij} - m_{kj}}{m_{kj}} |- \beta_1)) & \text{if } A_i > X_i \\
        0 & \text{otherwise}
        \end{cases} ,
\end{align}
where $F$ is a logistic sigmoid function and $\beta_0$ and $\beta_1$ corresponde to the advantage value and the probability that this particular advantage value will be enough to be preferred. For example, consider the case where $\beta_0 = 0.1$ and $\beta_1 = 50$. This would mean that the advantage of 10\% would be preferred with 50\% probability. Consequently, if the difference is 20\%, then it will be preferred with 99\% probability. The logistic function brings the notion of ``soft" comparison instead of ``hard" comparison in which case the small differences would be ignored, while large differences would be extremely preferred \citep{noguchi2018multialternative}.

@@

To effectively illustrate the functioning of MDbS, employing an example choice set that encapsulates the context effects discussed can be beneficial. Hence, example dataset depicted in figure \ref{fig:MDBsContextExample} will be used. Although there are five alternatives in the figure \ref{fig:MDBsContextExample}, only three of them will be discussed at a time.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{staticFiles/contextEffectExampleScatterplot.png}
    \caption{Example choice set to explain MDbS' account for big three context effects. $A$ and $B$ are considered original two alternatives (binary choice set). $D$ is dominated by $A$ on both dimensions. $C$ acts as compromise between the original two and $S$ is a similar option to $B$. Although these effects would be present in different variations of the choice set (for example one can make $A$ as compromise option), for the simplicity I will concentrate on this example.} % Add your description here
    \label{fig:MDBsContextExample} % This labels your figure for reference

@@

When adding option $D$ which is dominated by $A$ in both dimensions to binary choice set $A$ $B$ (hereafter binary choice set will be used instead of $A$ and $B$) one creates an attraction effect \citep{huberEtAl82, huberPuto83}. Huber et al. \citeyearonly{huberEtAl82} explained attraction effect via weight shifts for individuals. Addition of $D$ would make people to weigh dimension 2 more. Hence, $A$ and $D$ will have higher ``interest" among people, and this is where $A$ wins over $D$.

@@

In MDbS this is explained by $\beta_0$ and $\beta_1$ from equation \ref{probabilityOneIsFavored}. Recall that, the reason of having the sigmoid function with arguments $\beta_0$ and $\beta_1$ is to make sure that small differences would be relatively ignored. This is in line with previous literature stating that people tend to ignore small differences between alternatives when making a choice \citep{kalwani1992consumer}. Hence, the small differences between $B$ and $S$ would be ignored, which would be translated to decreased probability of any of them winning over the other when being compared. This will indirectly increase the evidence accumulation for $A$ while resulting in ``shared" evidence between $B$ and $S$.

Not only MDbS accounts for big three context effect, it also successfully accounts for other known effects to decision making literature such as: attribute spacing effect \citep{cooke1998multiattribute}, centrality effect \citep{brown2011decision}, background contrast effect \citep{tverskySimonson93}, endowment effect \citep{knetsch1989endowment} and etc.\footnote{For the full list of the effects MDbS can account for please refer to Noguchi \citeyearonly{noguchi2018multialternative}.} Overall, authors claim that MDbS can theoretically account for up to 25 context effect variations.

@@

After delving into the theoretical mechanisms of MDbS and its handling of the big three context effects, the focus now shifts to the practical component of the research: the methodology. The subsequent section provides a deeper exploration into the specifics of the research process. It offers a brief overview of the dataset, followed by an in-depth explanation of the steps involved in the analysis.

@@

@@

Attribute range effect which was first investigated by Mellers \citeyearonly{mellers1994trade} which described people's tendency to scale perceived attractiveness of an alternative on given attribute using the entire range of that attribute. Hence, $\beta_0$ in MDbS represents the fraction of the difference between attributes compared to the entire attribute range which is bounded between 0 and 1. On the other hand, $\beta_1$ represents the percentage of preference of that difference for an individual, hence it is also bounded with values between 0 and 100. I have created 99 $\beta_0$ values evenly spaced between 0 and 1 and 99 $\beta_1$ values evenly spaced between 0 and 100.

@@

\begin{algorithm}
\caption{Simplified Differential Evolution.}
\begin{algorithmic}[1]

@@

 $$\text{Similarity}(F) = O(\text{Cluster}_F) ,
$$
